Template for indicator documentation

### 2023-09-07 DRAFT_functional_plant_indicators_natopen.Rmd
corrections of some types and formatting issues. Nat-open is not yet finished (OK until heading "Analyses").

- as test data we use plant community data from (1) the ANO monitoring scheme (cf. Tingstad et al. 2019) and (2) from the GRUK monitoring scheme (cf. Evju et al. 2020).
- as reference data we use (3) generalized species lists developed by Evju et al. 2023 for nat-open ecosystem types.
- (4) Swedish plant indicator data for light, nitrogen, ans soil disturbance from Tyler et al. (2021), and (5) Grime's CSR values for plant species' strategies (towards competition, stress, and ruderal conditions)

@@

(2) GRUK monitoring data
GRUK stands for 'Grunnlendt åpen kalkmark', and represents a monitoring of open, limestone rich ecosystems with soils too shallow for forest establishment. The limestone rich focus of the scheme is motivated from a high biodiversity value and places all monitoring sites in an area around the Oslofjord. As of spring 2023, the GRUK data comprise 596 vegetation plots in 146 sites, all of which are mapped to the ecosystem type "åpen grunnlendt mark" (T2) and therein to the limestone-rich subtypes T2-C-7 and T2-C-8 in the Norwegian classification system for ecosystem types.
The vegetation analysis in GRUK record percent cover for every species of vascular plants in 1 sqm plots. In addition, a range of site variables related to ecosystem-specific drivers of state are recorded.
In the analysis in this document, we use all GRUK monitoring data data, which span the years 2020-22. The 2022-data also include an assessment of site condition, which is included in the analysis of results in this document. In the future, these data data may become available in Miljødirektoratets kartkatalog.

@@

(4) The Swedish plant indicator set published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we decided to include indicator data for light, moisture, pH, nitrogen, phosphorus, grazing_mowing, and soil disturbance for semi-natural ecosystems as these are thought to be subject to potential change due to abandonment, drainage/flooding, pollution, and erosion.

(5) Grime's system of plant strategy scores (Grime 1974) comprises relative (too one another) scores for the competition-, stress-, and disturbance("ruderality")-related life strategy of plant species. In the analysis, in this document we use all three variables, C, S and R, as different pressures acting on the ecosystem might change every one of the strategies (e.g. alien species for competition, climate change for stress, land use change for ruderality).

@@

## Reference state and values
### Reference state
The reference state is defined via the functional signature of the generalized species lists for NiN ecosystem types (see also Töpper et al. 2018). For the nat-open ecosystem types these lists have been newly prepared by Evju et al. 2023. By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator. These distributions are calculated for minor ecosystem types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the major ecosystem types (hovedtyper) in NiN. A more extensive discussion on the use of reference communities can be found in Jakobsson et al. (2020).

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

>>>>

- as test data we use plant community data from (1) the ANO monitoring scheme (cf. Tingstad et al. 2019) and (2) from the GRUK monitoring scheme (cf. Evju et al. 2020)
- as reference data we use (3) generalized species lists developed by Evju et al. (2023) for nat-open ecosystem types
- (4) Swedish plant indicator data for light, nitrogen, and soil disturbance from Tyler et al. (2021), and (5) Grime's CSR values for plant species' strategies (towards competition, stress, and ruderal conditions)

@@

(2) GRUK monitoring data
GRUK stands for 'Grunnlendt åpen kalkmark', and represents a monitoring of open, limestone rich ecosystems with soils too shallow for forest establishment. The limestone rich focus of the scheme is motivated from a high biodiversity value and places all monitoring sites in an area around the Oslofjord. As of spring 2023, the GRUK data comprise 596 vegetation plots in 146 sites, all of which are mapped to the ecosystem type "åpen grunnlendt mark" (T2) and therein to the limestone-rich subtypes T2-C-7 and T2-C-8 in the Norwegian classification system for ecosystem types.
The vegetation analysis in GRUK records percent cover for every species of vascular plants in 1 m^2 plots. In addition, a range of site variables related to ecosystem-specific drivers of state are recorded.
In the analysis in this document, we use all GRUK monitoring data, which span the years 2020-22. The 2022 data also include an assessment of site condition, which is included in the analysis of results in this document. In the future, these data data may become available in Miljødirektoratets kartkatalog.

@@

(4) The Swedish plant indicator set published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we decided to include indicator data for light, moisture, pH, nitrogen, phosphorus, grazing_mowing, and soil disturbance for semi-natural ecosystems, as these are thought to be subject to potential change due to abandonment, drainage/flooding, pollution, and erosion.

(5) Grime's system of plant strategy scores (Grime 1974) comprises relative (too one another) scores for the competition-, stress-, and disturbance("ruderality")-related life strategy of plant species. In the analysis in this document, we use all three variables, C, S and R, as different pressures acting on the ecosystem might change every one of the strategies (e.g. alien species for competition, climate change for stress, land use change for ruderality).

@@

## Reference state and values
### Reference state
The reference state is defined via the functional signature of the generalized species lists for NiN ecosystem types (see also Töpper et al. 2018). For the nat-open ecosystem types these lists have been newly prepared by Evju et al. (2023). By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator. These distributions are calculated for minor ecosystem types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the major ecosystem types (hovedtyper) in NiN. A more extensive discussion on the use of reference communities can be found in Jakobsson et al. (2020).

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Framstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

### 2023-09-07 DRAFT_functional_plant_indicators_seminat.Rmd
corrections of some types and formatting issues. Nat-open is not yet finished (OK until heading "Analyses").

```
In the ANO data deviations in the grazing/mowing index occur mainly in semi-natural grasslands (T32) on intermediate soils in terms of limestone (C3, C4, C6, C5 [but the latter is one category higher in limestone than intermediate). The deviatons towards low grazing/mowing values indicate that management and use are too extensive.

@@

```
In the ASO data the nitrogen deviations indicate too high nitrogen content in limestone richer (intermediate to very rich) semi-natural grasslands.

>>>>

```
In the ANO data, deviations in the grazing/mowing index occur mainly in semi-natural grasslands (T32) on intermediate soils in terms of limestone (C3, C4, C6, C5 [but the latter is one category higher in limestone than intermediate]). The deviations towards low grazing/mowing values indicate that management and use are too extensive.

@@

```
In the ASO data, the nitrogen deviations indicate too high nitrogen content in limestone richer (intermediate to very rich) semi-natural grasslands.
